---
title: "Stats in R - Day 1"
output: html_notebook
---

```{r loading_libraries, include=FALSE}
library(tidyverse)
```


# Exercise 1

Sampling from distributions in R is done using the `r<distrib>()` functions, where <distrib> is substituted for the short name of a distribution. Some distributions you might be exposed to (and their sampling functions) are:

* The Normal (Gaussian) distribution - `rnorm()`
* The Poisson distribution - `rpois()`
* The Gamma distribution - `rgamma()`
* The Exponential distribution - `rexp()`
* The Binomial distribution - `rbinom()`

Each of these functions are called with an argument `n` for the number of samples to take, as well as arguments that specify the parameters of the distribution. For example, a normal distribution is defined by it's mean and standard deviation, so these can be provided to `rnorm()`

```{r distribution_sampling}
#Using default mean (0) and standard deviation (1)
normal_default <- rnorm(n = 10)

#Setting mean and standard deviation
normal_specified <- rnorm(n = 10, mean = 5, sd = 3)

normal_default

normal_specified
```

Using these sampling functions, complete the following exercises. Some initial code has been provided but you will need to fill in the blanks (????? or _____) and extend them to other distributions.

>Sample from the Normal, Binomial and one other distribution.  Take 5, 10, 200 and 1000 samples with default parameters and histogram the results.   

```{r changing_sample_size}
#Normal distribution
norm_5 <- rnorm(n = 5)
norm_10 <- rnorm(n = 10)
norm_200 <- rnorm(_____)
_____

#Histograms
hist(norm_5)
hist(norm_10)
_____
_____
```

> Take 30 samples and repeat this 6 times for your chosen distributions. Histogram the observations. How does the histogram of the samples look different each time?

```{r repeated_sampling}
#Normal distribution
repeated_norm <- tibble(a = rnorm(30), b = rnorm(30), c = rnorm(30), d = rnorm(30), e = rnorm(30), f = rnorm(30))

repeated_norm

#Histograms with hist()
hist(repeated_norm$a)
hist(repeated_norm$b)
_____

#Histograms with ggplot
long_norm <- gather(repeated_norm, sample_id, sample_value)

long_norm

ggplot(long_norm, aes(x = sample_value)) + geom_histogram() + facet_wrap(~ sample_id)
```

> Create box and whisker plots for your repeated samples. Compare with the histograms, do they summarise the distributions well?

```{r boxplots}
#Boxplots with boxplot()
boxplot(repeated_norm)

#Boxplots with ggplot
ggplot(long_norm, aes(x=sample_id, y = sample_value)) + geom_?????()
```

> Using the inbuilt datasets `sleep` and `InsectSprays`. Explore their data with histograms and boxplots.

```{r real_data}
#Learn about the datasets
?sleep
?InsectSprays

#Histograms


#Boxplots

```

# Exercise 3E


